An efficient algorithm for integer lattice reduction

### 2023-08-13 tex/paper.tex
revising

\documentclass{article}
\usepackage[margin=1in]{geometry}
\usepackage{amsthm,amsmath,amssymb}
\usepackage{graphicx}
\usepackage{natbib}
\usepackage[hidelinks,breaklinks]{hyperref}
\pdfsuppresswarningpagegroup=1

\newtheorem{theorem}{Theorem}[]
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{corollary}[theorem]{Corollary}
\newtheorem{definition}[theorem]{Definition}
\newtheorem{observe}[theorem]{Observation}
\newtheorem{remark1}[theorem]{Remark}
\newenvironment{remark}{\begin{remark1} \rm}{\end{remark1}}

\DeclareMathOperator{\nint}{nint}
\DeclareMathOperator*{\argmin}{argmin}
\DeclareMathOperator{\bigoh}{\mathcal{O}}
\def\Id{\rm Id}

\title{An efficient algorithm for integer lattice reduction}
\author{Fran\c{c}ois Charton, Kristin Lauter, Cathy Li, and Mark Tygert}

@@

\begin{abstract}
A lattice of integers is the collection of all linear combinations
of a set of vectors for which all entries of the vectors are integers
and all coefficients in the linear combinations are also integers.
Lattice reduction refers to the problem of finding a set of vectors
in a given lattice such that the collection of all integer linear combinations
of this subset is still the entire original lattice and so that
the Euclidean norms of the subset are reduced.
The present paper proposes simple, efficient iterations for lattice reduction
which are guaranteed to reduce the Euclidean norms of the basis vectors
(the vectors in the subset) monotonically during every iteration.
Each iteration selects the basis vector for which projecting off
(with integer coefficients) the components of the other basis vectors
along the selected vector minimizes the Euclidean norms
of the reduced basis vectors.
Each iteration projects off the components along the selected basis vector
and efficiently updates all information required for the next iteration
to select its best basis vector and perform the associated projections.
\end{abstract}

@@

The set of $n$ vectors whose linear combinations form the lattice
is known as a ``basis'' for the lattice.
Traditionally the vectors in the basis are required to be linearly independent,
but throughout the present paper, the term ``basis'' will abuse terminology
slightly in omitting any requirement to be linearly independent.
The term ``reduced basis'' refers
to any unimodular integer linear transformation of a basis.
(A unimodular integer linear transformation is multiplication
with an invertible $n \times n$ matrix such that all entries of the matrix
and its inverse are integers; the entries of the inverse of a matrix
of integers are integers if and only if the absolute value
of the determinant of the matrix is 1.)
The goal of lattice reduction (the topic of the present paper)
is to minimize the Euclidean norms of the vectors in the reduced basis.

@@

The projection of a column vector $v$ onto a column vector $w$ is $c \cdot w$,
with the scalar coefficient $c$ defined to be the real number
%
\begin{equation}
c = \frac{v^\top w}{w^\top w},
\end{equation}
%
where $v^\top$ denotes the transpose of $v$
(and $w^\top$ denotes the transpose of $w$).
Needless to say, $v^\top w$ is the inner product between $v$ and $w$,
while $w^\top w$ is the inner product of $w$ with itself, which is of course
the square of the Euclidean norm of $w$.
Unfortunately, even if all entries of both $v$ and $w$ are integers,
subtracting off this projection $c \cdot w$ from $v$ directly
would result in a vector, $v - c \cdot w$, whose entries may not be integers.
Fortunately, all entries of the vector $v - \nint(c) \cdot w$
would remain integers, and Subsection~\ref{theory} below proves that
the Euclidean norm of $v - \nint(c) \cdot w$ is less than or equal to
the Euclidean norm of $v$. (Here and throughout the present paper,
$\nint(c)$ denotes the result of rounding $c$ to the nearest integer.)
In fact, if both $v$ and $w$ are in a lattice of integers,
then $v - \nint(c) \cdot w$ is in the same lattice of integers
and is shorter (or at least no longer) than $v$.
Theorem~\ref{theorem} below elaborates.

Let us denote by $a^0_1$, $a^0_2$, \dots, $a^0_n$ the initial basis vectors,
each being a column vector of $m$ integers.
The present paper proposes an iterative algorithm that reduces the basis
from iteration $i$ to iteration $i+1$ from $a^i_1$, $a^i_2$, \dots, $a^i_n$
to $a^{i+1}_1$, $a^{i+1}_2$, \dots, $a^{i+1}_n$, starting with $i=0$.
Each iteration $i$ of the algorithm selects an index $k$
that minimizes the Euclidean norms $\| a^i_j - c^i_{j,k} \cdot a^i_k \|$
for all $j \ne k$, where the scalar projection coefficient $c^i_{j,k}$ is
%
\begin{equation}
\label{rounded}
c^i_{j,k} = \nint\left( \frac{(a^i_j)^\top a^i_k}{(a^i_k)^\top (a^i_k)} \right)
\end{equation}
%
for $j \ne k$ and
%
\begin{equation}
c^i_{j,k} = 0
\end{equation}
%
for $j = k$.
Specifically, the index $k$ is chosen to minimize the sum of the $p$-th powers
of the Euclidean norms, that is, $k$ minimizes
$\sum_{j = 1}^n \| a^i_j - c^i_{j,k} \cdot a^i_k \|^p$.
(If $p = 2$, then this sum is the square of the Frobenius norm of the matrix.)
Iteration $i$ constructs the matrix for the next iteration as
$a^{i+1}_j = a^i_j - c^i_{j,k} \cdot a^i_k$.

{\it Prima facie}, considering all possible indices in order to minimize
the sum of the $p$-th powers appears to be computationally expensive.
However, the sum being minimized can be evaluated efficiently
via the Gram matrix whose entries are $g^i_{j,k} = (a^i_j)^\top a^i_k$,
as can the projection coefficients $c^i_{j,k}$ defined in~(\ref{rounded}).
Moreover, updating the entries of the Gram matrix from one iteration
to the next is also computationally efficient.

The algorithm of the present paper can complement others for lattice reduction,
such as the ``LLL'' algorithm introduced by~\cite{lenstra-lenstra-lovasz}.
Two entire books about the LLL algorithm are those
of~\cite{nguyen-vallee} and~\cite{bremner}.
The numerical examples presented below report the results
of running the algorithm of~\cite{lenstra-lenstra-lovasz}
in conjunction with the algorithm of the present paper.
The implementation of the LLL algorithm used here is
significantly more efficient (albeit less robust to worst-case,
adversarial examples devised for applications outside cryptography)
than the others' reviewed by~\cite{schnorr-euchner} and~\cite{stehle};
see Subsection~\ref{implementation} below
for details.\footnote{Permissively licensed open-source codes implementing
both the algorithm of the present paper and the classical LLL algorithm
of~\cite{lenstra-lenstra-lovasz} are available
at \url{https://github.com/facebookresearch/latticer}}

The primary purpose of the algorithm of the present paper is to polish
the outputs of other algorithms for lattice reduction;
on its own, the algorithm of the present paper tends to get stuck
in local minima, with the iterations attaining an equilibrium
that is far away from optimally minimizing the sum of the $p$-th powers.
The convergence is monotonic, but need not arrive at the optimal minimum
when fully converged.

The remainder of the present paper has the following structure:
Section~\ref{methods} describes and analyzes the algorithm in detail. 
Section~\ref{results} illustrates the algorithm via several numerical examples,
comparing and combining the proposed algorithm with the classic method
of~\cite{lenstra-lenstra-lovasz}.
Section~\ref{conclusion} draws some conclusions.
Appendix~\ref{poor} sketches several seemingly natural alternatives
that performed poorly in numerical experiments;
this appendix also points to other authors' variations of the LLL algorithm.
Appendix~\ref{further} supplements the figures of Section~\ref{results}
with further figures.

@@

This subsection describes the algorithm of the present paper in detail.

Consider the $n$ column vectors $a^0_1$, $a^0_2$, \dots, $a^0_n$,
each of size $m \times 1$.
The proposed scheme makes no assumption on the relative sizes of $m$ and $n$
--- $m$ can be less than, equal to, or greater than $n$.
Calculate the entries of the symmetric square Gram matrix 
%
\begin{equation}
\label{Gram0}
g^0_{j,k} = (a^0_j)^\top (a^0_k)
\end{equation}
%
for $j = 1$, $2$, \dots, $n$, and $k = 1$, $2$, \dots, $n$,
where $(a^0_j)^\top$ denotes the transpose of $a^0_j$.
(Of course, $(a^0_j)^\top (a^0_k)$ is simply the inner product
between $a^0_j$ and $a^0_k$.)
This costs $\bigoh(mn^2)$ operations.

Iterations, $i = 0$, $1$, $2$, \dots, will maintain the relation
%
\begin{equation}
\label{Gram}
g^i_{j,k} = (a^i_j)^\top (a^i_k)
\end{equation}
%
for $j = 1$, $2$, \dots, $n$, and $k = 1$, $2$, \dots, $n$.

@@

Calculate the projection coefficients
%
\begin{equation}
c^i_{k,k} = 0
\end{equation}
%
for $k = 1$, $2$, \dots, $n$, and
%
\begin{equation}
\label{coeffs}
c^i_{j,k} = \nint\left( \frac{(a^i_j)^\top a^i_k}
                             {(a^i_k)^\top (a^i_k)} \right)
          = \nint\left( \frac{g^i_{j,k}}{g^i_{k,k}} \right)
\end{equation}
%
for $j = 1$, $2$, \dots, $n$, and $k = 1$, $2$, \dots, $n$ with $k \ne j$,
where $\nint$ denotes the nearest integer.
This costs $\bigoh(n^2)$ operations.

The sum of the $p$-th powers of the Euclidean norms
when projecting off the $k$-th vector is
%
\begin{equation}
\label{sos}
s^i_k = \sum_{j=1}^n \| a^i_j - c^i_{j,k} \, a^i_k \|^p
= \sum_{j=1}^n \left( g^i_{j,j} + (c^i_{j,k})^2 \, g^i_{k,k}
- 2 c^i_{j,k} \, g^i_{j,k} \right)^{p/2}
\end{equation}
%
for $k = 1$, $2$, \dots, $n$.
The iterations are likely to work better by starting with $p = 2$
and only later running with, say, $p = 1$.
This costs $\bigoh(n^2)$ to compute.

@@

Project off the $\tilde{\imath}$-th vector to update every vector
%
\begin{equation}
\label{ortho}
(a^{i+1}_k)_j
= (a^i_k)_j - c^i_{k,\tilde{\imath}} \, (a^i_{\tilde{\imath}})_j
\end{equation}
%
for $j = 1$, $2$, \dots, $m$, and $k = 1$, $2$, \dots, $n$,
where $(a^i_k)_j$ denotes the $j$-th entry of $a^i_k$.
This costs $\bigoh(mn)$.

Update the Gram matrix via the relation
%
\begin{equation}
\label{Gram_update}
g^{i+1}_{j,k}
= (a^{i+1}_j)^\top (a^{i+1}_k)
= (a^i_j - c^i_{j,\tilde{\imath}} \, a^i_{\tilde{\imath}})^\top
  (a^i_k - c^i_{k,\tilde{\imath}} \, a^i_{\tilde{\imath}})
= g^i_{j,k}
+ c^i_{j,\tilde{\imath}} \, c^i_{k,\tilde{\imath}}
\, g^i_{\tilde{\imath},\tilde{\imath}}
- c^i_{j,\tilde{\imath}} \, g^i_{\tilde{\imath},k}
- c^i_{k,\tilde{\imath}} \, g^i_{j,\tilde{\imath}}
\end{equation}
%
for $j = 1$, $2$, \dots, $n$, and $k = 1$, $2$, \dots, $n$.
This costs $\bigoh(n^2)$.

@@

The purpose of this subsection is to state and prove Theorem~\ref{theorem},
elaborating the facts stated in the fourth paragraph of the introduction,
Section~\ref{intro}.

@@

\begin{proof}
Since the sign of $\nint(r)$ is the same as the sign of $r$,
the sign of the left-hand side of~(\ref{simplified}) is the sign of $r$
times the sign of
%
\begin{equation}
\label{final}
\nint(r) - 2 r.
\end{equation}
%
Now, the sign of~(\ref{final}) is opposite to the sign of $r$:
if $r > 1/2$, then $\nint(r) - 2r \le r+1/2-2r = 1/2-r < 0$;
if $r < -1/2$, then $\nint(r) - 2r \ge r-1/2-2r = -1/2-r > 0$;
if $|r| < 1/2$, then $\nint(r) - 2r = -2r$, whose sign is opposite to $r$'s.
Combining these observations yields~(\ref{simplified}).
\end{proof}

@@

The following theorem states that projection using the coefficients
defined in~(\ref{coeffs}) never increases the Euclidean norm.
This implies that the above algorithm never increases the Euclidean norm
of any of the basis vectors at any time; the sum of the $p$-th powers
of the Euclidean norms therefore converges monotonically to a local minimum,
for every possible (positive) value of $p$ simultaneously.
%
\begin{theorem}
\label{theorem}
Suppose that $c^i_{j,k}$ is the coefficient defined in~(\ref{coeffs})
for $i = 0$, $1$, $2$, \dots, $j = 1$, $2$, \dots, $n$,
and $k = 1$, $2$, \dots, $n$.
Then
%
\begin{equation}
\label{mono}
\| a^i_j - c^i_{j,k} \, a^i_k \| \le \| a^i_j \|
\end{equation}
%
for $i = 0$, $1$, $2$, \dots, $j = 1$, $2$, \dots, $n$,
and $k = 1$, $2$, \dots, $n$.
\end{theorem}

\begin{proof}
The square of the right-hand side of~(\ref{mono}) is
%
\begin{equation}
\label{rhs}
\| a^i_j \|^2 = g^i_{j,j}.
\end{equation}
%
The square of the left-hand side of~(\ref{mono}) is
%
\begin{equation}
\label{lhs}
\| a^i_j - c^i_{j,k} \, a^i_k \|^2
= g^i_{j,j} + (c^i_{j,k})^2 \, g^i_{k,k} - 2 c^i_{j,k} \, g^i_{j,k}.
\end{equation}
%
Formula~(\ref{update}) shows that~(\ref{lhs})
is less than or equal to~(\ref{rhs}).
\end{proof}

@@

This section presents the results
of several numerical experiments.\footnote{Permissively licensed open-source
software that automatically reproduces all results reported
in the present paper is available at
\url{https://github.com/facebookresearch/latticer}}
Subsection~\ref{figures} describes the figures.
Subsection~\ref{examples} constructs the examples whose numerical results
the figures report.
Subsection~\ref{implementation} provides details of the implementation
and computer system used.
Subsection~\ref{discussion} discusses the empirical results.

@@

This subsection describes empirical results, illustrating the algorithms
via Figures~\ref{p2time1-1e-15}--\ref{p2err1-1e-15-31}.
Supplementary figures are available in Appendix~\ref{further}.

@@

The figures report on two kinds of experiments.
The first kind runs the LLL algorithm followed by ours,
once in each of ten trials. Each of the ten trials permutes the basis vectors
at random, with a different random permutation. The points plotted
in the figures are the means of the ten trials. The plotted bars range
from the associated minimum over the ten trials to the associated maximum
(not displaying the standard deviation that error bars sometimes report).
The figures refer to this single randomized permutation followed by LLL
followed by ours as ``run only once.''

The second kind of experiment runs the following sequence ten times
in succession: (1) random permutation of the basis vectors
followed by (2) LLL followed by (3) the algorithm of the present paper,
with each of the first nine times feeding its output into the input
of the next time. The figures refer to this sequence of ten
as ``run repeatedly.''

The figures report the reduction either in the Frobenius norm
of the matrix of basis vectors or in the minimum of the Euclidean norms
of the basis vectors. (The Frobenius norm is the square root of the sum
of the squares of the entries of the matrix.)
The figures report the reduction due to LLL run on its own,
as well as the further reduction due to post-processing
the output from LLL via the algorithm of the present paper.
Therefore, the full reduction in norm is the product
of the reported fractions remaining after reduction,
as all runs polish the outputs of the LLL algorithm via ours.
The fraction reduced by ours that is reported for the sequence run repeatedly
pertains only to the very last run of ours,
whereas the fraction reduced by LLL reported
for the same sequence run repeatedly pertains to the entire series of ten runs
of LLL followed by ours, aside from separating out the very last run of ours.

@@

Appendix~\ref{further} presents further figures,
reporting results for different values of $\delta$ and $p$.
The conclusions that can be drawn from the further figures
appear to be broadly consistent with those presented in the present section.

@@

>>>>

% SIAM Article Template
\documentclass[hidelinks,onefignum,onetabnum]{siamart220329}

% Information that is shared between the article and the supplement
% (title and author information, macros, packages, etc.) goes into
% ex_shared.tex. If there is no supplement, this file can be included
% directly.

\input{shared}
\usepackage[algoruled]{algorithm2e}
%%%%\SetAlgoSkip{smallskip}
%%%%\SetAlgoInsideSkip{smallskip}
%%%%\SetAlCapFnt{\small}
%%%%\SetAlCapNameFnt{\small}
%%%%\setlength{\interspacetitleruled}{\smallskipamount}

% Optional PDF information
\ifpdf
\hypersetup{
  pdftitle={An efficient algorithm for integer lattice reduction},
  pdfauthor={Fran\c{c}ois Charton, Kristin Lauter, Cathy Li, and Mark Tygert}
}
\fi

% The next statement enables references to information in the
% supplement. See the xr-hyperref package for details.

\externaldocument[][nocite]{supplement}

% FundRef data to be entered by SIAM
%<funding-group specific-use="FundRef">
%<award-group>
%<funding-source>
%<named-content content-type="funder-name"> 
%</named-content> 
%<named-content content-type="funder-identifier"> 
%</named-content>
%</funding-source>
%<award-id> </award-id>
%</award-group>
%</funding-group>

@@

% REQUIRED
\begin{abstract}
A lattice of integers is the collection of all linear combinations
of a set of vectors for which all entries of the vectors are integers
and all coefficients in the linear combinations are also integers.
Lattice reduction refers to the problem of finding a set of vectors
in a given lattice such that the collection of all integer linear combinations
of this subset is still the entire original lattice and so that
the Euclidean norms of the subset are reduced.
The present paper proposes simple, efficient iterations for lattice reduction
which are guaranteed to reduce the Euclidean norms of the basis vectors
(the vectors in the subset) monotonically during every iteration.
Each iteration selects the basis vector for which projecting off
(with integer coefficients) the components of the other basis vectors
along the selected vector minimizes the Euclidean norms
of the reduced basis vectors.
Each iteration projects off the components along the selected basis vector
and efficiently updates all information required for the next iteration
to select its best basis vector and perform the associated projections.
\end{abstract}

% REQUIRED
\begin{keywords}
matrix, projection, orthogonalization, optimization, cryptography,
number theory
\end{keywords}

% REQUIRED
\begin{MSCcodes}
65F25, 65K10, 11T71
\end{MSCcodes}

@@

The set of $n$ vectors whose linear combinations form the lattice
is known as a ``basis'' for the lattice.
Traditionally the vectors in the basis are required to be linearly independent,
but throughout the present paper, the term ``basis'' will abuse terminology
slightly in omitting any requirement to be linearly independent.
The term ``reduced basis'' refers
to any unimodular integer linear transformation of a basis.
(A unimodular integer linear transformation is multiplication
with an invertible $n \times n$ matrix such that all entries of the matrix
and its inverse are integers; the entries of the inverse of a matrix
of integers are integers if and only if the absolute value
of the determinant of the matrix is 1 --- this follows straightforwardly
from the Cramer Rule for solving systems of linear equations.)
The goal of lattice reduction (the topic of the present paper)
is to minimize the Euclidean norms of the vectors in the reduced basis.

@@

The projection of a column vector $v$ onto a column vector $w$ is $c \cdot w$,
with the scalar coefficient $c$ defined to be the real number
%
\begin{equation}
c = \frac{w^\top v}{w^\top w},
\end{equation}
%
where $w^\top$ denotes the transpose of $w$.
Needless to say, $w^\top v$ is the inner product between $w$ and $v$,
while $w^\top w$ is the inner product of $w$ with itself, which is of course
the square of the Euclidean norm of $w$.

Unfortunately, even if all entries of both $v$ and $w$ are integers,
subtracting off this projection $c \cdot w$ from $v$ directly
would result in a vector, $v - c \cdot w$, whose entries may not be integers.
Fortunately, all entries of the vector $v - \nint(c) \cdot w$
would remain integers, and we prove in Subsection~\ref{theory} below that
the Euclidean norm of $v - \nint(c) \cdot w$ is less than or equal to
the Euclidean norm of $v$. (Here and throughout the present paper,
$\nint(c)$ denotes the result of rounding $c$ to the nearest integer.
The reader is welcome to use any convention desired for the definition
of $\nint(k + 1/2)$, where $k$ is an integer, provided that the convention
is used consistently. The proofs in Subsection~\ref{theory} below omit
special consideration of these special cases, as the proofs are trivial
in such cases.) In fact, if both $v$ and $w$ are in a lattice of integers,
then $v - \nint(c) \cdot w$ is in the same lattice of integers
and is shorter (or at least no longer) than $v$.
Theorem~\ref{theorem} in Subsection~\ref{theory} below elaborates.
Mapping the pair $v$ and $w$ to the pair $v - \nint(c) \cdot w$ and $w$
is clearly a unimodular integer linear transformation,
as is any composition of such mappings
(since the product of unimodular integer linear transformations
is just another unimodular integer linear transformation).

Let us denote by $a^0_1$, $a^0_2$, \dots, $a^0_n$ the initial basis vectors,
each being a column vector of $m$ integers.
The present paper proposes an iterative algorithm that reduces the basis
from iteration $i$ to iteration $i+1$ from $a^i_1$, $a^i_2$, \dots, $a^i_n$
to $a^{i+1}_1$, $a^{i+1}_2$, \dots, $a^{i+1}_n$, starting with $i=0$.
Each iteration $i$ of the algorithm selects an index $k$
that minimizes the Euclidean norms $\| a^i_j - c^i_{j,k} \cdot a^i_k \|$
for all $j \ne k$, where the scalar projection coefficient $c^i_{j,k}$ is
%
\begin{equation}
\label{rounded}
c^i_{j,k} = \nint\left( \frac{(a^i_k)^\top a^i_j}{(a^i_k)^\top (a^i_k)} \right)
\end{equation}
%
for $j \ne k$ and
%
\begin{equation}
c^i_{j,k} = 0
\end{equation}
%
for $j = k$.
Specifically, the index $k$ is chosen to minimize the sum of the $p$-th powers
of the Euclidean norms, that is, $k$ minimizes
$\sum_{j = 1}^n \| a^i_j - c^i_{j,k} \cdot a^i_k \|^p$.
(If $p = 2$, then this sum is the square of the Frobenius norm of the matrix.)
Iteration $i$ constructs the matrix for the next iteration as the matrix whose
columns are the basis vectors $a^{i+1}_j = a^i_j - c^i_{j,k} \cdot a^i_k$.

{\it Prima facie}, considering all possible indices in order to minimize
the sum of the $p$-th powers appears to be computationally expensive.
However, the sum being minimized can be evaluated efficiently
via the Gram matrix whose entries are $g^i_{j,k} = (a^i_k)^\top a^i_j$,
as can the projection coefficients $c^i_{j,k}$ defined in~(\ref{rounded}).
Moreover, updating the entries of the Gram matrix from one iteration
to the next is also computationally efficient.

The algorithm of the present paper can complement others for lattice reduction,
such as the ``LLL'' algorithm introduced by~\cite{lenstra-lenstra-lovasz}.
Two entire books about the LLL algorithm are those
of~\cite{nguyen-vallee} and~\cite{bremner}.
The numerical examples presented in Section~\ref{results} below report
the results of running the algorithm of~\cite{lenstra-lenstra-lovasz}
in conjunction with the algorithm of the present paper.
The implementation of the LLL algorithm used here is
significantly more efficient (albeit less robust to worst-case,
adversarial examples devised for applications outside cryptography)
than the others' reviewed by~\cite{schnorr-euchner} and~\cite{stehle};
see Subsection~\ref{implementation} below
for details.\footnote{Permissively licensed open-source codes implementing
both the algorithm of the present paper and the classical LLL algorithm
of~\cite{lenstra-lenstra-lovasz} are available
at \url{https://github.com/facebookresearch/latticer}}

The primary purpose of the algorithm of the present paper is to polish
the outputs of other algorithms for lattice reduction;
on its own, the algorithm of the present paper tends to get stuck
in local minima, with the iterations attaining a fixed-point equilibrium
that is far away from optimally minimizing the sum of the $p$-th powers.
The convergence is monotonic, but need not arrive at the optimal minimum
when fully converged.

The remainder of the present paper has the following structure:
Section~\ref{methods} describes and analyzes the algorithm in detail. 
Section~\ref{results} illustrates the algorithm via several numerical examples,
comparing and combining the proposed algorithm with the classic method
of~\cite{lenstra-lenstra-lovasz}.
Section~\ref{conclusion} draws some conclusions.
Section~\ref{poor} of the Supplementary Materials
sketches several seemingly natural alternatives
that performed poorly in numerical experiments;
this supplementary section also points to other authors' variations
of the LLL algorithm.
Section~\ref{further} of the Supplementary Materials complements the figures
of Section~\ref{results} with further figures.

@@

This subsection describes the algorithm of the present paper in detail.
Pseudocode is available in Algorithm~\ref{ours}.

Consider the $n$ column vectors $a^0_1$, $a^0_2$, \dots, $a^0_n$,
each of size $m \times 1$.
The proposed scheme makes no assumption on the relative sizes of $m$ and $n$;
that is, $m$ can be less than, equal to, or greater than $n$.
Calculate the entries of the symmetric square Gram matrix 
%
\begin{equation}
\label{Gram0}
g^0_{j,k} = (a^0_k)^\top (a^0_j)
\end{equation}
%
for $j = 1$, $2$, \dots, $n$, and $k = 1$, $2$, \dots, $n$,
where $(a^0_k)^\top$ denotes the transpose of $a^0_k$.
(Of course, $(a^0_k)^\top (a^0_j)$ is simply the inner product
between $a^0_k$ and $a^0_j$.)
This costs $\bigoh(mn^2)$ operations.

Iterations, $i = 0$, $1$, $2$, \dots, will maintain the relation
%
\begin{equation}
\label{Gram}
g^i_{j,k} = (a^i_k)^\top (a^i_j)
\end{equation}
%
for $j = 1$, $2$, \dots, $n$, and $k = 1$, $2$, \dots, $n$.

@@

Calculate the projection coefficients
%
\begin{equation}
c^i_{k,k} = 0
\end{equation}
%
for $k = 1$, $2$, \dots, $n$, and
%
\begin{equation}
\label{coeffs}
c^i_{j,k} = \nint\left( \frac{(a^i_k)^\top a^i_j}
                             {(a^i_k)^\top (a^i_k)} \right)
          = \nint\left( \frac{g^i_{j,k}}{g^i_{k,k}} \right)
\end{equation}
%
for $j = 1$, $2$, \dots, $n$, and $k = 1$, $2$, \dots, $n$ with $k \ne j$,
where $\nint$ denotes the nearest integer.
This costs $\bigoh(n^2)$ operations.

The sum of the $p$-th powers of the Euclidean norms
when projecting off the $k$-th vector is
%
\begin{equation}
\label{sos}
s^i_k = \sum_{j=1}^n \| a^i_j - c^i_{j,k} \, a^i_k \|^p
= \sum_{j=1}^n \left( g^i_{j,j} + (c^i_{j,k})^2 \, g^i_{k,k}
- 2 c^i_{j,k} \, g^i_{j,k} \right)^{p/2}
\end{equation}
%
for $k = 1$, $2$, \dots, $n$.
This costs $\bigoh(n^2)$ to compute.

@@

Project off the $\tilde{\imath}$-th vector to update every vector
%
\begin{equation}
\label{ortho0}
a^{i+1}_k
= a^i_k - c^i_{k,\tilde{\imath}} \, a^i_{\tilde{\imath}}
\end{equation}
%
for $k = 1$, $2$, \dots, $n$, that is,
%
\begin{equation}
\label{ortho}
(a^{i+1}_k)_j
= (a^i_k)_j - c^i_{k,\tilde{\imath}} \, (a^i_{\tilde{\imath}})_j
\end{equation}
%
for $j = 1$, $2$, \dots, $m$, and $k = 1$, $2$, \dots, $n$,
where $(a^i_k)_j$ denotes the $j$-th entry of $a^i_k$.
This costs $\bigoh(mn)$.

Update the Gram matrix via the relation
%
\begin{equation}
\label{Gram_update}
g^{i+1}_{j,k}
= (a^{i+1}_k)^\top (a^{i+1}_j)
= (a^i_k - c^i_{k,\tilde{\imath}} \, a^i_{\tilde{\imath}})^\top
  (a^i_j - c^i_{j,\tilde{\imath}} \, a^i_{\tilde{\imath}})
= g^i_{j,k}
+ c^i_{j,\tilde{\imath}} \, c^i_{k,\tilde{\imath}}
\, g^i_{\tilde{\imath},\tilde{\imath}}
- c^i_{j,\tilde{\imath}} \, g^i_{\tilde{\imath},k}
- c^i_{k,\tilde{\imath}} \, g^i_{j,\tilde{\imath}}
\end{equation}
%
for $j = 1$, $2$, \dots, $n$, and $k = 1$, $2$, \dots, $n$.
This costs $\bigoh(n^2)$.

@@

The purpose of this subsection is to state and prove Theorem~\ref{theorem},
elaborating the facts stated in the fifth paragraph of the introduction,
Section~\ref{intro}.

@@

\begin{proof}
Since the sign of $\nint(r)$ is the same as the sign of $r$ (or 0),
the sign of the left-hand side of~(\ref{simplified}) is the sign of $r$ (or 0)
times the sign of
%
\begin{equation}
\label{final}
\nint(r) - 2 r.
\end{equation}
%
Now, the sign of~(\ref{final}) is opposite to the sign of $r$:
if $r > 1/2$, then $\nint(r) - 2r \le r+1/2-2r = 1/2-r < 0$;
if $r < -1/2$, then $\nint(r) - 2r \ge r-1/2-2r = -1/2-r > 0$;
if $|r| < 1/2$, then $\nint(r) - 2r = -2r$, whose sign is opposite to $r$'s.
Combining these observations yields~(\ref{simplified}).
\end{proof}

@@

The following theorem states that projection using the coefficients
defined in~(\ref{coeffs}) never increases the Euclidean norm.
This implies that the above algorithm never increases the Euclidean norm
of any of the basis vectors at any time; the sum of the $p$-th powers
of the Euclidean norms therefore converges monotonically to a local minimum,
for every possible (positive) value of $p$ simultaneously.
%
\begin{theorem}
\label{theorem}
Suppose that $c^i_{j,k}$ is the coefficient defined in~(\ref{coeffs})
for $i = 0$, $1$, $2$, \dots, $j = 1$, $2$, \dots, $n$,
and $k = 1$, $2$, \dots, $n$.
Then the Euclidean norms satisfy
%
\begin{equation}
\label{mono}
\| a^i_j - c^i_{j,k} \, a^i_k \| \le \| a^i_j \|
\end{equation}
%
for $i = 0$, $1$, $2$, \dots, $j = 1$, $2$, \dots, $n$,
and $k = 1$, $2$, \dots, $n$.
\end{theorem}

\begin{proof}
According to~(\ref{Gram}), the square of the right-hand side of~(\ref{mono}) is
%
\begin{equation}
\label{rhs}
\| a^i_j \|^2 = g^i_{j,j}.
\end{equation}
%
The square of the left-hand side of~(\ref{mono}) is
%
\begin{equation}
\label{lhs}
\| a^i_j - c^i_{j,k} \, a^i_k \|^2
= g^i_{j,j} + (c^i_{j,k})^2 \, g^i_{k,k} - 2 c^i_{j,k} \, g^i_{j,k}.
\end{equation}
%
Formula~(\ref{update}) shows that~(\ref{lhs})
is less than or equal to~(\ref{rhs}).
\end{proof}

@@

This section presents the results
of several numerical experiments.\footnote{Permissively licensed open-source
software that automatically reproduces all results reported
in the present paper is available at
\url{https://github.com/facebookresearch/latticer}}
Subsection~\ref{figures} describes the figures.
Subsection~\ref{examples} constructs the examples whose numerical results
the figures report.
Subsection~\ref{implementation} provides details of the implementation
and computer system used.
Subsection~\ref{simple} describes the tables for a dead-simple example included
due to a reviewer's request.
Subsection~\ref{discussion} discusses the empirical results.

@@

This subsection describes empirical results, illustrating the algorithms
via Figures~\ref{p2time1-1e-15}--\ref{p2err1-1e-15-31}.
Supplementary figures are available in Section~\ref{further}
of the Supplementary Materials.

@@

The figures report on two kinds of experiments.
The first kind runs the LLL algorithm followed by ours,
once in each of ten trials. Each of the ten trials permutes the basis vectors
at random, with a different random permutation.
(The algorithm of the present paper is invariant to the ordering
of the basis vectors, whereas the LLL algorithm is sensitive to the ordering.)
The points plotted in the figures are the means of the ten trials.
So, for example, the times reported in Figure~\ref{p2time1-1e-15}
are averages over ten trials.
The plotted bars range from the associated minimum over the ten trials
to the associated maximum (not displaying the standard deviation
that error bars sometimes report).
Figures~\ref{p2err1-1e-15-13once} and~\ref{p2err1-1e-15-31once}
include such bars ranging from the minimum to the maximum of the ten trials.
The figures refer to this single randomized permutation followed by LLL
followed by ours as ``run once.''
Figures~\ref{p2time1-1e-15}, \ref{p2err1-1e-15-13once},
and~\ref{p2err1-1e-15-31once} concern this first kind of experiment.

The second kind of experiment runs the following sequence ten times
in succession: (1) random permutation of the basis vectors
followed by (2) LLL followed by (3) the algorithm of the present paper,
with each of the first nine times feeding its output into the input
of the next time. Algorithm~\ref{combination} formalizes this series
of ten three-step sequences explicitly. The figures refer to the sequence
of ten as ``run repeatedly.'' Figures~\ref{p2err1-1e-15-13}
and~\ref{p2err1-1e-15-31} concern this second kind of experiment.
The first steps (random permutation) affect the LLL algorithm,
which is sensitive to the ordering of the basis vectors
(while the algorithm of the present paper is invariant to the ordering).
The points plotted in Figures~\ref{p2err1-1e-15-13} and~\ref{p2err1-1e-15-31}
are the final results of the series of ten three-step sequences
(whereas the points plotted in Figures~\ref{p2time1-1e-15},
\ref{p2err1-1e-15-13once}, and~\ref{p2err1-1e-15-31once}
are the means over ten independent trials).

Figure~\ref{p2time1-1e-15} displays the time taken per trial.
Figures~\ref{p2err1-1e-15-13once}--\ref{p2err1-1e-15-31} report
the reduction either in the Frobenius norm
of the matrix of basis vectors or in the minimum of the Euclidean norms
of the basis vectors. (The Frobenius norm is the square root of the sum
of the squares of the entries of the matrix.)
The figures report the reduction due to LLL run on its own,
as well as the further reduction due to post-processing
the output from LLL via the algorithm of the present paper.
Therefore, the full reduction in norm is the product
of the reported fractions remaining after reduction,
as all runs polish the outputs of the LLL algorithm via ours.
The fraction reduced by ours that is reported for the sequence run repeatedly
pertains only to the very last run of ours,
whereas the fraction reduced by LLL reported
for the same sequence run repeatedly pertains to the entire series of ten runs
of LLL followed by ours, aside from separating out the very last run of ours.
The plots displaying the fraction reduced by the last run of ours
that is reported for the sequence run repeatedly feature titles,
``reduction in Frob.\ norm by ours after reps.\ of ours+LLL''
and ``reduction in min.\ norm by ours after reps.\ of ours+LLL,''
abbreviating ``repetitions'' to ``reps.''
In all the figures, the plots on the right display the results
of running the algorithm of Section~\ref{methods}
(that is, Algorithm~\ref{ours}) after having run either the LLL algorithm
or all of Algorithm~\ref{combination} except for the final run
of the algorithm of Section~\ref{methods} (which is Algorithm~\ref{ours}).

@@

Section~\ref{further} of the Supplementary Materials presents further figures,
reporting results for different values of $\delta$ and $p$.
The conclusions that can be drawn from the further figures
appear to be broadly consistent with those presented in the present section.

@@

\subsection{Description of the tables}
\label{simple}
